Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-21 paper/paper.tex

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
not in the scope of \pu.  However, this does not cause any imprecision
in the way  of the GC marking more  cells as live, since \pu\  is yet to
come into existence at $\pi_7$  and its liveness will not be consulted
by the garbage collector at this point.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================

In an eager language, \pw\ would not have been live after $\pi_7$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
every program  point from $\pi_8$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

@@

>>>>

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
not in the scope of \pu.  However, this does not cause any imprecision
in the way  of the GC marking more  cells as live, since \pu\  is yet to
come into existence at $\pi_7$  and its liveness will not be consulted
by the garbage collector at this point.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================
In an eager language, \pw\ would not have been live after $\pi_7$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
every program  point from $\pi_8$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

@@

%\comment{

\section{Generating Liveness Equations}
\begin{figure*}[t!]
\begin{picture}(100,130)(0,-90)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.5\textwidth}
      {\sf
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\pi_3\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_1~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_2~(1~+~\py))$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}              (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \py) \IN~$\pi_{9}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.5\textwidth}
      {\sf
        \begin{tabular}{lc}
        $\cal{M} = $ 
            \begin{tabular}{lc}            
              $[$
              $\pi_{\mbox{7}} \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $\pi_{\mbox{4}} \mapsto \lbrace \rbrace , $\\
              $\pi_{\mbox{3}} \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P} = [$
              \begin{tabular}{lc}
              $\pi_{\mbox{1}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \
              $ \pi_{\mbox{2}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \\
              $ \pi_{\mbox{3}} \mapsto \lbrace \pi_{\mbox{3}} \rbrace, $ \
              $ \pi_{\mbox{4}} \mapsto \lbrace \pi_{\mbox{4}} \rbrace, $ \
              $ \pi_{\mbox{5}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \\
              $ \pi_{\mbox{6}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \
              $ \pi_{\mbox{7}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace $ \\
              $]$
            \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex

\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

 For our example program which has  a
single function \length, this generates the following equations:

  \begin{eqnarray*}
    \Lf{\length}{1}{\sigma}
    &=& \epsilonset \cup \acdr\Lf{\length}{1}{\epsilon}
   \end{eqnarray*}

\subsection{Generating Liveness Equations for Function bodies}
For the running example, $\length$ has calls from $\mainpgm$ at $\pi_8$
and a recursive call at $\pi_6$.
\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
So $\sigma_{\length} =
     \deltacall{\length}{\pi_8}{\mainpgm}  \cup \deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:

\begin{align*}
\sigma_{\length}    &=
 \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{\length}}}
\end{align*}

For the running example, containing unique GC points (corresponding to unique {\em e-paths} sets) $\pi_3, \pi_4, \pi_7$ in $\length$ and
$\pi_9$ in $e_\mainpgm$, this gives

\begin{align*}
\Lanv{7}{\xl} &= \acdr\Lf{\length}{1}{\epsilon} \cup \epsilonset \\
\Lanv{7}{\pu} &= \Lf{\length}{1}{\epsilon} \\
\Lanv{7}{\py} &= \epsilonset \\
\Lanv{3}{\xl} &= \epsilonset \\
\Lanv{3}{\px} &= \epsilonset 
\\[1ex]
  \Lanv{9}{\pb}  &= \Lf{\length}{1}{\sigma_{\mathit{\!all}}} \\
  \Lanv{9}{\pw}  &= \sigma_{\mathit{\!all}}
\end{align*}

\subsection{Solving liveness equations---the grammar interpretation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr\rbrace$.  Let  \var{$X$} denote the non-terminal for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{7}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{7}{\xl}}  &\rightarrow&   \var{\acdr \Lf
  {\length}{1}{\epsilon}} \mid \epsilon \\
\end{eqnarray*}

%}

% \item
